\chapter{Checklists, Templates, and Further Reading}
\section{Chapter Overview}
The appendix collects reusable assets that keep teams aligned.

\begin{lstlisting}[caption={Representing checklist ownership},label={lst:appendix_overview}]
from __future__ import annotations


def checklist_owner(name: str) -> str:
    """Map checklist names to owners."""
    return f"{name} owner: platform team"
\end{lstlisting}

\section{Launch Checklist}
Before shipping, confirm that formatting, linting, typing, and tests pass locally and in CI.
Review documentation updates, rotate credentials if needed, and run deployment rehearsals.

\begin{lstlisting}[caption={Evaluating launch readiness programmatically},label={lst:appendix_launch}]
from __future__ import annotations


def launch_ready(checks: dict[str, bool]) -> bool:
    """Return True when every checklist item passes."""
    return all(checks.values())
\end{lstlisting}

\section{Template Repository}
Maintain a starter repo containing \texttt{pyproject.toml}, \texttt{noxfile.py}, CI
workflows, and documentation scaffolding.  New services can fork it to inherit best
practices.

\begin{lstlisting}[caption={Scaffolding a new repository from a template},label={lst:appendix_template}]
from __future__ import annotations

import shutil
from pathlib import Path


def scaffold_from_template(template: Path, destination: Path) -> None:
    """Copy the template repo into a new project directory."""
    shutil.copytree(template, destination, dirs_exist_ok=True)
\end{lstlisting}

\section{Further Reading}
Recommended resources include Brett Slatkin's \emph{Effective Python}, Hynek Schlawack's
\emph{Solid Python}, the official \emph{Python Packaging User Guide}, and the \emph{Twelve
Factor App}.  Follow tool maintainers (Black, Ruff, Pytest, Mypy) to stay current.

\begin{lstlisting}[caption={Generating reading lists},label={lst:appendix_reading}]
from __future__ import annotations


def reading_plan(topics: list[str]) -> list[str]:
    """Pair topics with recommended books."""
    return [f"{topic}: Effective Python (2nd Ed.)" for topic in topics]
\end{lstlisting}

\section{Summary}
Standardised checklists, templates, and curated resources ensure every new initiative starts
with the same proven guardrails.

\begin{lstlisting}[caption={Summarising appendix assets},label={lst:appendix_summary}]
from __future__ import annotations


def summarize_assets(counts: dict[str, int]) -> str:
    """Provide a short statement listing asset counts."""
    return ", ".join(f"{kind}: {count}" for kind, count in counts.items())
\end{lstlisting}

\section*{Exercises}
\begin{enumerate}
  \item Build a personalised launch checklist and compare it with this appendix.
  \item Create a template repository for your organisation and solicit feedback.
  \item Curate a reading list tailored to your team's domain (web, data, ML).
  \item Run a brown-bag session where each engineer shares one tool configuration tip.
  \item Evaluate whether your onboarding guide references every relevant checklist; update
  as needed.
  \item Extend Listing~\ref{lst:appendix_launch} so it prints which checklist item failed
  when the launch is blocked.
\end{enumerate}
